Modeling the impacts of replacing car trips with combined public transport and cycling: \ Reproducible methods, results and actionable evidence from biclaR

### 2024-05-06 paper/PaperCEUS/PaperCEUS_rev1.Rmd
map municipalities plot
something not right regarding the percentages

---
title: "Modeling the impacts of replacing car trips with combined public transport and cycling: \ Reproducible methods, results and actionable evidence from biclaR"
author:
  - name: Rosa Félix\corref{cor1}
    affiliation: CERIS
    correspondingauthor: true
    # footnote: 1
    email: rosamfelix@tecnico.ulisboa.pt
  - name: Filipe Moura
    affiliation: CERIS
    email: fmoura@tecnico.ulisboa.pt
  - name: Robin Lovelace
    affiliation: ITS
    email: r.lovelace@uleeds.uk
address:
  - code: CERIS
    address: "CERIS, Instituto Superior Técnico - Universidade de Lisboa, Av Rovisco Pais 1, 1049-001 Lisboa, Portugal"
  - code: ITS
    address: "Institute for Transport Studies, University of Leeds. 34-40 University Rd, Leeds LS2 9JT, UK"
abstract: |
  A high proportion of car trips can be replaced by a combination of public transit and cycling for the first-and-last mile. This paper estimates the potential for cycling combined with public transit as a substitute for car trips in the Lisbon metropolitan area and assesses its socio-environmental impacts using open data and open source tools.
  A decision support tool that facilitates the design and development of a metropolitan cycling network was developed (_biclaR_). The social and environmental impacts were assessed using the _HEAT for Cycling_ and the _HEAT as a Service_ tools. The impacts of shifting car trips to public transport were also estimated and monetized.
  The results show that 10% of all trips could be made by cycling in combination with public transport. Shifting to cycling for the shorter first and last mile stages can reduce annual CO~2~eq emissions from 3,000 to 7,500 tons/day, while for the public transport leg, the transfer from car avoids of up to 20,500 tons of CO~2~eq emissions per year. The estimated socio-environmental benefits are of €125 million to €325 million over 10 years.
  This evidence can support policymakers to prioritize interventions that reduce the reliance on private motor vehicles.
  \
keywords: 
  - Active transport
  - Intermodality
  - First and last mile
  - Health economic assessment
  - Environmental impacts
  - Open data and methods
journal: "Computers, Environment and Urban Studies"
date: "`r Sys.Date()`"
classoption: review, doubleblind, 3p, authoryear #change later to preprint instead of review
bibliography: bibliography.bib
biblio-style: elsarticle-harv
linenumbers: true
numbersections: true
# Use a CSL with `citation_package = "default"`
csl: https://www.zotero.org/styles/computers-environment-and-urban-systems
# csl: computers-environment-and-urban-systems.csl
preamble: |
  \biboptions{authoryear}
  \usepackage{url}
  \usepackage{float}
  \usepackage{booktabs}
  \usepackage{longtable}
  \usepackage{makecell}
  \usepackage{multirow}
output: 
  rticles::elsevier_article:
    keep_tex: yes
    # citation_package: "default" #ou nenhum
    latex_engine: xelatex
    extra_dependencies: "subfig"
#   bookdown::word_document2:
#     fig_caption: true
# always_allow_html: true
urlcolor: blue
citecolor: black
linkcolor: blue
# natbiboptions: longnamesfirst,angle,semicolon
---
\hypersetup{citecolor=black}

@@

>>>>

---
title: "Modeling the impacts of replacing car trips with combined public transport and cycling: \ Reproducible methods, results and actionable evidence from biclaR"
author:
  - name: Rosa Félix\corref{cor1}
    affiliation: CERIS
    correspondingauthor: true
    # footnote: 1
    email: rosamfelix@tecnico.ulisboa.pt
  - name: Filipe Moura
    affiliation: CERIS
    email: fmoura@tecnico.ulisboa.pt
  - name: Robin Lovelace
    affiliation: ITS
    email: r.lovelace@uleeds.uk
address:
  - code: CERIS
    address: "CERIS, Instituto Superior Técnico - Universidade de Lisboa. Av Rovisco Pais 1, 1049-001 Lisboa, Portugal"
  - code: ITS
    address: "Institute for Transport Studies, University of Leeds. 34-40 University Rd, Leeds LS2 9JT, UK"
abstract: |
  A high proportion of car trips can be replaced by a combination of public transit and cycling for the first-and-last mile. This paper estimates the potential for cycling combined with public transit as a substitute for car trips in the Lisbon metropolitan area and assesses its socio-environmental impacts using open data and open source tools.
  A decision support tool that facilitates the design and development of a metropolitan cycling network was developed (_biclaR_). The social and environmental impacts were assessed using the _HEAT for Cycling_ and the _HEAT as a Service_ tools. The impacts of shifting car trips to public transport were also estimated and monetized.
  The results show that 10% of all trips could be made by cycling in combination with public transport. Shifting to cycling for the shorter first and last mile stages can reduce annual CO~2~eq emissions from 3,000 to 7,500 tons/day, while for the public transport leg, the transfer from car avoids of up to 20,500 tons of CO~2~eq emissions per year. The estimated socio-environmental benefits are of €125 million to €325 million over 10 years.
  This evidence can support policymakers to prioritize interventions that reduce the reliance on private motor vehicles.
  \
keywords: 
  - Active transport
  - Intermodality
  - First and last mile
  - Health economic assessment
  - Environmental impacts
  - Open data and methods
journal: "Computers, Environment and Urban Studies"
date: "`r Sys.Date()`"
classoption: review, doubleblind, 3p, authoryear #change later to preprint instead of review
bibliography: bibliography.bib
biblio-style: elsarticle-harv
linenumbers: true
numbersections: true
# Use a CSL with `citation_package = "default"`
csl: https://www.zotero.org/styles/computers-environment-and-urban-systems
# csl: computers-environment-and-urban-systems.csl
preamble: |
  \biboptions{authoryear}
  \usepackage{url}
  \usepackage{float}
  \usepackage{booktabs}
  \usepackage{longtable}
  \usepackage{makecell}
  \usepackage{multirow}
output: 
  rticles::elsevier_article:
    keep_tex: yes
    # citation_package: "default" #ou nenhum
    latex_engine: xelatex
    extra_dependencies: "subfig"
#   bookdown::word_document2:
#     fig_caption: true
# always_allow_html: true
urlcolor: blue
citecolor: black
linkcolor: blue
# natbiboptions: longnamesfirst,angle,semicolon
---
\hypersetup{citecolor=black}

@@

<!-- INSERT ZONAL MAP HERE -->

```{r mapzones, fig.cap="Potential Bike + PT daily trips for the 10% target, by municipality"}
CAR_PT_TRANSF = readRDS("load/CAR_TP_TRANSF.Rds")

# summarize by OD pair
CAR_PT_TRANSF_ODs = CAR_PT_TRANSF |>
  filter(LTS == 4) |> 
  group_by(DICOFREor11, DICOFREde11) |>
  summarise(Total = sum(Total),
            Car = sum(Car),
            Bike = sum(Bike),
            Walk = sum(Walk),
            TP = sum(TP),
            Bike_new4 = sum(Bike_new4),
            Bike_new10 = sum(Bike_new10),
            Bike_4 = sum(Bike_4),
            Bike_10 = sum(Bike_10),
            TP_4 = sum(TP_4),
            TP_10 = sum(TP_10)) |>
  mutate(Bikeper = 100*Bike/Total,
         Bike_new4per = 100*Bike_new4/Total,
         Bike_new10per = 100*Bike_new10/Total,
         TPper = 100*TP/Total,
         TP_4per = 100*TP_4/Total,
         TP_10per = 100*TP_10/Total) |> 
  ungroup()

# summarize by origin
CAR_PT_TRANSF_mun_or = CAR_PT_TRANSF_ODs |> 
  group_by(DICOFREor11) |>
  summarise(Total = sum(Total),
            Car = sum(Car),
            Bike = sum(Bike),
            Walk = sum(Walk),
            TP = sum(TP),
            Bike_new4 = sum(Bike_new4),
            Bike_new10 = sum(Bike_new10),
            Bike_4 = sum(Bike_4),
            Bike_10 = sum(Bike_10),
            TP_4 = sum(TP_4),
            TP_10 = sum(TP_10)) |>
  mutate(Bikeper = 100*Bike/Total,
         Bike_new4per = 100*Bike_new4/Total,
         Bike_new10per = 100*Bike_new10/Total,
         TPper = 100*TP/Total,
         TP_4per = 100*TP_4/Total,
         TP_10per = 100*TP_10) |> 
  ungroup()

# join with municipalities geo
CAR_PT_TRANSF_mun_or_geo = CAR_PT_TRANSF_mun_or |> 
  left_join(FREGUESIASgeo_zones |> select(Dicofre, Concelho) |> sf::st_drop_geometry(),
            by = c("DICOFREor11" = "Dicofre"))
CAR_PT_TRANSF_mun_or_geo$Concelho[CAR_PT_TRANSF_mun_or_geo$Concelho == "Setúbal"] = "Setubal"
CAR_PT_TRANSF_mun_or_geo = CAR_PT_TRANSF_mun_or_geo |> 
  group_by(Concelho) |>
  summarise(Total = sum(Total),
            Car = sum(Car),
            Bike = sum(Bike),
            Walk = sum(Walk),
            TP = sum(TP),
            Bike_new4 = sum(Bike_new4),
            Bike_new10 = sum(Bike_new10),
            Bike_4 = sum(Bike_4),
            Bike_10 = sum(Bike_10),
            TP_4 = sum(TP_4),
            TP_10 = sum(TP_10)) |>
  mutate(Bikeper = 100*Bike/Total,
         Bike_new4per = 100*Bike_new4/Total,
         Bike_new10per = 100*Bike_new10/Total,
         TPper = 100*TP/Total,
         TP_4per = 100*TP_4/Total,
         TP_10per = 100*TP_10/Total) |>
  left_join(MUNICIPIOSgeo, by = "Concelho") |> 
  ungroup()

# bind municipalities with zero potential
CAR_PT_TRANSF_mun_or_geo_18 = CAR_PT_TRANSF_mun_or_geo |> 
  dplyr::bind_rows(as.data.frame(
   MUNICIPIOSgeo |> filter(!Concelho %in% CAR_PT_TRANSF_mun_or_geo$Concelho))) |> 
  sf::st_as_sf()

# re-calculating the percentage of trips
CAR_PT_TRANSF_mun_or_geo_18 = CAR_PT_TRANSF_mun_or_geo_18 |> 
  left_join(TRIPSall_mun, by = c("Concelho" = "Origem")) |>

  #CONTINUE FROM HERE ---  LOOK AT DIFRERENCES TOTAL AND VIAGENS

mutate(Car_4 = Car - Bike_new4,
         Car_10 = Car - Bike_new10,
         TP_4 = TP + Bike_new4,
         TP_10 = TP + Bike_new10,)

# plot 1
ggplot(data = CAR_PT_TRANSF_mun_or_geo_18) +
  geom_sf(aes(fill = Bike_new10)) +
  viridis::scale_fill_viridis(option = "G", direction = -1, name = "Bike + PT trips", na.value = "white") +
  theme_void()
  # theme(legend.position = "bottom") +
  # labs(title = "Potential Bike + PT for 10% target, by municipality")

# plot 2
ggplot(data = CAR_PT_TRANSF_mun_or_geo_18) +
  geom_sf(aes(fill = TP_10per)) +
  viridis::scale_fill_viridis(option = "G", direction = -1, name = "Bike + PT trips %", na.value = "white") +
  theme_void()
  # theme(legend.position = "bottom") +
  # labs(title = "Potential Bike + PT for 10% target, by municipality")
